# Preface {.unnumbered}


## The moiraine package

Omics datasets provide an overview of the content of cells for a specific molecular layer (e.g. transcriptome, proteome, metabolome). By integrating different omics datasets obtained on the same biological samples, we can gain a deeper understanding of the interactions between these molecular layers, and shed light on the regulations occurring both within and between layers. A number of statistical methods have been developed to extract such information from multi-omics datasets, and many have been implemented in software such as R packages. However, these tools differ conceptually, in terms of the input data they require, the assumptions they make, the statistical approaches they use or even the questions they answers. They also differ at a practical level in terms of the format required for data input, the parameters to tune or select, and the format in which the results are returned. These differences render the **application of several integration tools to a multi-omics dataset** and the **comparison of their results** complex and time-consuming.

The `moiraine` package aims at alleviating these issues, by providing **a framework to easily and consistently apply different integration tools to a same multi-omics dataset**. It implements numerous visualisation and reporting functions to **facilitate the interpretation of the integration results** as well as the comparison of these results across integration methods. In addition, in an effort to make these computations reproducible, `moiraine` heavily relies on the [`targets` package](https://books.ropensci.org/targets/) for the **construction of reproducible analysis pipelines**.

## About this manual

In this manual, we are showcasing the functionalities of the `moiraine` package by presenting an in-depth walk-through example of a multi-omics integration analysis. This will not only talk about the how in terms of R functions etc, but also talk about the integration methods and how to use them.

Here, say that we heavily recommend to be familiar with targets (teaching targets it out of the scope of this manual, and we refer to the excellent targets manual).

```{r}
#| child: "_setup.qmd"
```

```{r loading-packages}
library(targets)
```

```{r}
1 + 1
```

A targets chunck:

```{targets}
tar_target(a, 2+2)
```

```{file}
"something"
```


