# Preface {.unnumbered}

Quick blurb around multi-omics integration. There are many tools available to perform multi-omics integration, and a lot are implemented as R packages. These tools differ conceptually (in terms of required data input, assumptions, questions they answer) but also at a practical level in terms of input data format, parameters, etc and output format. That makes it time-consuming to apply different tools to a same multi-omics dataset, and to compare the results.

## The moiraine package

The moiraine package aims at alleviating this by providing a framework to easily and consistently apply different integration tools to a same dataset. It also facilitates the comparison of results with consistent formatting of integration output and visualisations.

In addition, in an effort to make these computations reproducible, moiraine heavily relies on targets for the creating of reproducible pipelines.

## About this manual

In this manual, we are showcasing the functionalities of the moiraine package by presenting an in-depth walk-through example of multi-omics integration analysis. This will not only talk about the how in terms of R functions etc, but also talk about the integration methods and how to use them.

Here, say that we heavily recommend to be familiar with targets (teaching targets it out of the scope of this manual, and we refer to the excellent targets manual).

```{r}
#| child: "_setup.qmd"
```

```{r loading-packages}
library(targets)
```

This is a Quarto book.

To learn more about Quarto books visit <https://quarto.org/docs/books>.

```{r}
1 + 1
```

A targets chunck:

```{targets}
tar_target(a, 2+2)
```

```{file}
"something"
```

## 
